---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
# load the packages used
require(lubridate)
require(data.table)

unzip("activity.zip")

data = read.csv("activity.csv")
data$date = ymd(data$date) # convert the date to a data class

```

## What is mean total number of steps taken per day?

What is the total number of steps per day, with the days:

```{r}
tapply(data$steps,data$date,sum, na.omit = TRUE)
```


Plot a histogram of the number of steps (i.e. distribution)
```{r}
hist(tapply(data$steps,data$date,sum, na.omit = TRUE),xlab = "Num Steps", main = "Histogram of total number of steps per day")
```

What is the mean number of steps per day?
```{r}
mean_steps = mean(tapply(data$steps,data$date,sum, na.rm = TRUE), na.rm = TRUE)
```

The mean number of steps per day is: `r mean_steps`

what is the median number of steps per day?
```{r}
median_steps = median(tapply(data$steps,data$date,sum, na.rm = TRUE), na.rm = TRUE)
```

The median number of steps per day is: `r median_steps`

What is the average daily activity pattern?
```{r}
plot(tapply(data$steps, data$interval, mean, na.rm = TRUE), type = 'l', ylab = "Number of Steps", xlab = "Interval")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

The full interval with the maximum number of steps can be found using which.max as follows:
```{r, echo = TRUE}
data[which.max(data$steps),]
```


## Imputing missing values

What is the total number of missing data?
```{r}
sum(!complete.cases(data))
```


Devise a strategy for filling in the missing values:
1. Create a mean data frame
2. Iterate through the existing data frame, replacing NA values with the mean values

```{r}
interval_steps <- aggregate(steps ~ interval, data, mean)
for (i in 1:nrow(data)){
  if (is.na(data$steps[i])){
    interval_val <- data$interval[i]
    row_id <- which(interval_steps$interval == interval_val)
    steps_val <- interval_steps$steps[row_id]
    data$steps[i] <- steps_val
  }
}
```

the impact of this is:
```{r}
new_median_steps = median(tapply(data$steps,data$date,sum, na.rm = TRUE), na.rm = TRUE)
new_mean_steps = mean(tapply(data$steps,data$date,sum, na.rm = TRUE), na.rm = TRUE)
```

The old mean: `r mean_steps`  
The old median: `r median_steps`  

The new mean: `r new_mean_steps`  
the new median: `r new_median_steps`  

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor weekday vs weekend:

```{r}
day_of_week = 1:nrow(data)
data$day = wday(data$date)

 for (i in 1:nrow(data)){
  if (data[i,4] == 1 | data[i,4] == 7){
    day_of_week[i] = "Weekend"
  } else { day_of_week[i] = "weekday"}
}
data$day_of_week = day_of_week
data$day_of_week = factor(data$day_of_week)
stepsByDay <- aggregate(steps ~ interval + day_of_week, data = data, mean)
names(stepsByDay) <- c("interval", "day_of_week", "steps")

summary(stepsByDay)
head(stepsByDay)

```


Plot the data by weekend vs weekday:

```{r}
require(lattice)
xyplot(steps ~ interval | day_of_week, stepsByDay, type = "l")
```